---
title: "DV Assignment 2, Part 1: School-level differences in secondary school
advice for similar test performance"
author: "Group 4: Sander van Nieuwenhuijzen, Natasha Medved, Sophie Preysing"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))
library(lintr)
lintr::lint("DV-Assignment2-Part1-GroupX.Rmd")
```

## Research question
 "Do schools with similar test results give similar secondary-school
 advice?"
 
This research questions takes on an inferential approach. We analyse whether schools secondary-school advice is statistically related to school's average test results. Through the use of an appropriate statistical model, we aim to find out if a significant relation between these variables exist. Our target audience consists of policymakers who need information on national patterns and the magnitude of the potential difference between schools. Therefore, our final data visualization will present statistical evidence for the relationship between test results and secondary school advice on a national level. The information will be presented in a way that is easy to digest for someone without a background in data science. The results can provide relevant information for public debate, development of policies and scientific progress in the field of education.  

## Exploratory process
In the first exploratory plot we check whether schools advise higher or lower than their test results predict. Here, we compare its actual average advice against expected advice based on test results. 

In the second plot we explore trend lines of the subjects language, maths and reading. We plot schoolweging against the percentage of students reaching target level and check the test results per schoolweging, per subject.

Lastly, we make a box plot of providers against schoolweging. The plot is visualized below here, we explore whether some providers have a more disadvantaged school population. A high score for schoolweging indicates a more disadvantaged school population. We ordered the schools from lowest median schoolweging (left) to highest median schoolweging (right). We learned that provider "DOE" has the lowest median schoolweging and provider "DIA" the highest. However, the medians for different providers fluctuate very little. This suggests that there is no significant difference in schoolweging between providers. 

```{r exploratory-plot}
source("../scripts/03-clean-data.R") # creates school_rq3
viridis_colours <- viridisLite::viridis(10) # colourblind proof scheme

# shared theme, so all plots look alike
theme_explore <- theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))

plot_3 <- ggplot(school_rq3, aes(
  x = reorder(provider, schoolweging, FUN = median),
  y = schoolweging,
  fill = provider
)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  scale_fill_manual(values = viridis_colours) +
  scale_colour_manual(values = viridis_colours) +
  labs(
    title = "School disadvantage by primary test provider",
    subtitle = "Higher schoolweging indicates a more disadvantaged
    student population",
    x = "Primary test provider",
    y = "Schoolweging"
  ) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  theme_explore

plot_3
```

---

