---
title: "DV Assignment 2: [Your Title Here]"
author: "Group 4: Natasha medved, Sophie Preysing, Sander van Nieuwenhuijzen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))
```

## Research question

*What are you investigating, and why does it matter? Who is the target audience for your final visualization, and how does that shape your choices? You can start from what you wrote in Part 1.*

This research questions takes on an inferential approach. We analyse whether schools secondary-school advice is statistically related to school's average test results. Through the use of an appropriate statistical model, we aim to find out if a significant relation between these variables exist. Our target audience consists of policymakers who need information on national patterns and the magnitude of the potential difference between schools. Therefore, our final data visualization will present statistical evidence for the relationship between test results and secondary school advice on a national level. The information will be presented in a way that is easy to digest for someone without a background in data science. The results can provide relevant information for public debate, development of policies and scientific progress in the field of education.  


## Exploratory process

*Describe how you explored the data before settling on your final visualization. Include at least one exploratory plot below (this can be the one from Part 1, or an updated one), and a short note on what it told you and how it influenced your final visualization.*

Lastly, we make a box plot of providers against schoolweging. The plot is visualized below here, we explore whether some providers have a more disadvantaged school population. A high score for schoolweging indicates a more disadvantaged school population. We ordered the schools from lowest median schoolweging (left) to highest median schoolweging (right). We learned that provider "DOE" has the lowest median schoolweging and provider "DIA" the highest. However, the medians for different providers fluctuate very little. This suggests that there is no significant difference in schoolweging between providers. This means that providers apply similarly disadvantaged populations. If the advices outcomes still differ a lot between providers, it means the test structure produces differences. 

```{r exploratory-plot}
```

## Final visualization

*Your final, explanatory visualization goes below. Include a short write-up: what does it communicate, why is this the best way to communicate it, and how robust is the finding?*

```{r final-plot}
```

## Conclusion

*What can you conclude from your visualization? What are the limitations of this analysis?*

---

**Before submitting:** run the lintr check on this file, and re-knit from a clean environment (`rm(list = ls())`, already done above) to make sure everything reproduces from scratch.
