---
title: "Same Test Results, Different Advice"
subtitle: "DV Assignment 2 – Part 2 (Final report)"
author: "Group 4: Natasha, Sander van Nieuwenhuijzen, Sophie Preysing"
date: "October 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

library(tidyverse)
library(here)
```

```{r data}
# Load the cleaned data (created by scripts/03-clean-data.R).
# If the processed file does not exist yet, build it first.
data_path <- here("data", "processed", "school_rq3.rds")
if (!file.exists(data_path)) {
  source(here("scripts", "03-clean-data.R"))
}
school_rq3 <- readRDS(data_path)

# Centre score and schoolweging; LIB (largest provider) as reference
model_data <- school_rq3 |>
  mutate(
    score_c = score - mean(score),
    schoolweging_c = schoolweging - mean(schoolweging),
    provider = relevel(factor(provider), ref = "LIB")
  )
```

```{r model}
# Weighted by n_advised: percentages of larger schools are more reliable
rq3_model <- lm(
  pct_havo_plus ~ score_c * schoolweging_c + provider,
  data = model_data,
  weights = n_advised
)

# Predicted lines at the 10th, 50th and 90th percentile of schoolweging,
# for a LIB school (the most common provider)
sw_levels <- quantile(model_data$schoolweging, c(0.1, 0.5, 0.9))

pred_grid <- expand_grid(
  score = seq(min(model_data$score), max(model_data$score),
              length.out = 100),
  schoolweging = sw_levels
) |>
  mutate(
    score_c = score - mean(model_data$score),
    schoolweging_c = schoolweging - mean(model_data$schoolweging),
    provider = factor("LIB", levels = levels(model_data$provider))
  )

pred_grid <- bind_cols(
  pred_grid,
  as_tibble(predict(rq3_model, newdata = pred_grid, interval = "confidence"))
)

# Size of the gap at an average score (10th vs 90th percentile)
gap_data <- tibble(schoolweging = unname(sw_levels[c(1, 3)])) |>
  mutate(
    score = mean(model_data$score),
    score_c = 0,
    schoolweging_c = schoolweging - mean(model_data$schoolweging),
    provider = factor("LIB", levels = levels(model_data$provider))
  )
gap_data$fit <- predict(rq3_model, newdata = gap_data)

gap_size <- round(gap_data$fit[1] - gap_data$fit[2], 1)
```

# Research question and target audience

The doorstroomtoets (the mandatory Dutch primary school transition test for Group 8 pupils) has sparked numerous debates about equal opportunities in education. This motivated government officials to launch a new doorstroomtoets (transfer test) in 2023 (Current Debates and Reforms, 2026), with the hope that a newly devised test would improve access to higher education for pupils from less privileged backgrounds. It is therefore important to understand whether, under this new test, schools with similar test results give similar secondary-school advice, regardless of how disadvantaged their pupils are.

This project adopts an inferential approach. Rather than only describing how advice differs between schools, the analysis assesses whether schools with similar test results advise fewer pupils to HAVO or higher when their pupils are more disadvantaged (H1), and whether this gap grows as test results increase (H2).

The target audience is education policymakers, such as staff at the Ministry of Education (OCW). They are familiar with the debate about fair school advice and with the doorstroomtoets, but are not expected to read a regression output. The visualization shows that, even after introducing an improved doorstroomtoets, schools with more disadvantaged pupils advise fewer pupils to HAVO or higher than schools with the same test results. This signals a need to monitor school advice relative to test results, with attention to schools with high schoolweging.

# Data and preparation

Using the DUO doorstroomtoets data, this analysis combines all four DUO files for school year 2024-2025: schooladviezen (secondary-school advice), referentieniveaus (pupils per reference level in maths, reading and language), eindscores (test provider) and schoolweging (disadvantage score). One row in these files is one school location, not one pupil, so each file was first aggregated to one row per school and the four files were then joined.

Only regular primary schools were kept; special primary schools (Sbo) were excluded as these do not have test results. The outcome is the percentage of pupils with a HAVO, HAVO/VWO or VWO advice, out of all pupils with an advice on the PRO-VWO scale (VSO and "no advice possible" were excluded). This outcome is referred to as HAVO or higher. Test results are the average percentage of pupils reaching the target level (1S for maths, 2F for reading and language). The three subjects correlated moderately at school level (r = 0.57—0.68), so it was appropriate to combine them into one average which gave a more reliable picture of a school's overall performance than a single subject. Each school's test provider is the test taken by most of its pupils. Schools missing from any of the four files were removed, as were schools with fewer than 10 advised pupils because in such small groups, a single pupil shifts the percentage by more than 10 points.

DUO hides counts below 5 as "<5". Dropping every school with such a count would remove most of the data, so each "<5" was replaced with 2.5, which is the midpoint of 1—4. Because this is an assumption, the analysis was rerun with 1 and 4. >>>> add findings from robustness check

# Exploratory process

<!-- TODO: required by the brief. Add at least one exploratory graph
(e.g. the one from Part 1, or an updated version) and describe what it
showed and how it shaped the final visualization. -->

[Add exploratory graph and description]

```{r exploratory-plot, eval = FALSE}
# TODO: paste the exploratory plot code here and remove eval = FALSE
```

# Final explanatory visualization

```{r final-plot, fig.width = 11, fig.height = 7.5, out.width = "100%"}
line_labels2 <- pred_grid |>
  filter(score == max(score)) |>
  mutate(
    label = c(
      "Advantaged\nschools",
      "Average\nschools",
      "Disadvantaged\nschools"
    )[match(schoolweging, sw_levels)]
  )

final_plot2 <- ggplot() +
  # Schools (all providers)
  geom_point(
    data = model_data,
    aes(x = score, y = pct_havo_plus, colour = schoolweging),
    size = 1.2, alpha = 0.2
  ) +
  # Uncertainty bands of the lines
  geom_ribbon(
    data = pred_grid,
    aes(
      x = score, ymin = lwr, ymax = upr,
      fill = schoolweging, group = schoolweging
    ),
    alpha = 0.4
  ) +
  # Dark outline under each line
  geom_line(
    data = pred_grid,
    aes(x = score, y = fit, group = schoolweging),
    colour = "grey20", linewidth = 2
  ) +
  # Expected advice at three levels of disadvantage
  geom_line(
    data = pred_grid,
    aes(x = score, y = fit, colour = schoolweging, group = schoolweging),
    linewidth = 1.3
  ) +
  # Line labels
  geom_text(
    data = line_labels2,
    aes(x = score, y = fit, label = label, colour = schoolweging),
    hjust = 0, nudge_x = 1, size = 4.2, lineheight = 0.9,
    fontface = "bold"
  ) +
  # Gap arrow: dark outline, then orange arrow on top
  annotate(
    "segment",
    x = mean(model_data$score), xend = mean(model_data$score),
    y = gap_data$fit[1], yend = gap_data$fit[2],
    arrow = arrow(ends = "last", length = unit(0.29, "cm"), type = "closed"),
    colour = "grey15", linewidth = 1.4
  ) +
  annotate(
    "segment",
    x = mean(model_data$score), xend = mean(model_data$score),
    y = gap_data$fit[1], yend = gap_data$fit[2],
    arrow = arrow(ends = "last", length = unit(0.25, "cm"), type = "closed"),
    colour = "#FF8C00", linewidth = 1
  ) +
  annotate(
    "label",
    x = mean(model_data$score), y = max(gap_data$fit) + 3,
    label = paste0(
      "Same test results:\n", round(gap_size), " percentage points\n",
      "fewer pupils advised\nHAVO or higher"
    ),
    hjust = 1.05, vjust = 0, size = 4.5, linewidth = 0,
    fill = alpha("white", 0.7), fontface = "bold"
  ) +
  scale_colour_viridis_c(
    name = "Schoolweging\n(higher = more\ndisadvantaged)",
    breaks = c(20, 25, 30, 35),
    direction = -1, end = 0.9,
    guide = guide_colourbar(barheight = unit(5, "cm"))
  ) +
  scale_fill_viridis_c(direction = -1, end = 0.9, guide = "none") +
  scale_x_continuous(
    breaks = seq(0, 100, by = 25),
    expand = expansion(mult = c(0.02, 0.3))
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  coord_cartesian(clip = "off") +
  labs(
    title = paste(
      "At the same test results, schools with more disadvantaged pupils",
      "advise\nfewer pupils to HAVO or higher"
    ),
    subtitle = paste0(
      "Lines show the average share of pupils advised HAVO or higher for ",
      "advantaged, average and disadvantaged\nschools, estimated by a ",
      "regression model. The gap grows as test results rise."
    ),
    x = "Test results: % of pupils reaching the target level",
    y = "% of pupils advised HAVO or higher",
    caption = paste0(
      "Dots: schools (all test providers). Test results: average % of ",
      "pupils reaching the target level in maths, reading and language. ",
      "Lines: regression\npredictions at the 10th, 50th and 90th ",
      "percentile of schoolweging, controlling for test provider (shown ",
      "for LIB, the most common test); shaded bands:\n95% confidence ",
      "intervals. N = ", format(nrow(model_data), big.mark = ","),
      " regular primary schools with 10+ advised pupils, 2024-2025; ",
      "'<5' counts replaced by 2.5. Source: DUO."
    )
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 14, colour = "grey25"),
    plot.caption = element_text(
      size = 9.5, colour = "grey40", hjust = 0, lineheight = 1.15
    ),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    panel.grid.minor = element_blank()
  )

final_plot2
```

Each dot is one of the 5,718 schools, placed by its test result (x-axis) and its share of pupils advised HAVO-or-higher (y-axis), and coloured by schoolweighing from yellow-green (advantage) to dark purple (disadvantage). The three lines show the model's predicted advice for schools at the 10th, 50th, and 90th percentile of schoolweging, with shaded 95% confidence intervals. The orange arrow marks the main finding: at an average test result, a disadvantaged school advises 17 percentage points fewer pupils to HAVO or higher than an advantaged school. The lines also fan out to the right. The gap is small in low-scoring schools and grows to around 30 percentage points among the highest-scoring ones.

The lines come from a linear regression of the percentage of HAVO-or-higher advice on test results, schoolweging and their interaction, controlling for test provider and weighted by the number of advised pupils (R^2^ = 0.66). Test results and scholweging were centred, so each main effect applies to a school with average values. At average test results, each extra point of schoolweging goes with 1.76 percentage points less HAVO-or-higher advice (95% CI - 1.84, p < 0.001): higher test results translate less into HAVO-or-higher advice in more disadvantaged schools, which supports H2.

# Conclusion and robustness

Schools with similar test results do not give similar advice. At the same test result, schools with more disadvantaged pupils advise fewer pupils to HAVO or higher. For two schools with an average test result on the same test, the disadvantaged scholl (90th percentile of schoolweging) advises 17 percentage points fewer pupils to HAVO or higher than the advantaged school (10th percentile). The gap grows as test results rise, so it is largest for disadvantaged schools whose pupils score well. For policy makers, this indicates that first, pupils in disadvantaged schools may be under-advised compared with similar pupils elsewhere, yet this is exactly the inequality the 2023 doorstroomtoets was meant to reduce. Second, the gap is largest for high-scoring schools which means the pupils most likely to miss out on access to higher education are the ones who perform well at disadvantaged schools.

```{r robustness}
# Each robustness model changes one choice; the schoolweging effect (H1)
# and the interaction (H2) should keep their sign and significance.

# Maths only, instead of the average of three subjects
rob1_model_data <- model_data |>
  mutate(maths_c = pct_maths_1s - mean(pct_maths_1s))

rob1_model <- lm(
  pct_havo_plus ~ maths_c * schoolweging_c + provider,
  data = rob1_model_data,
  weights = n_advised
)

# Average advice on the 1-6 scale, instead of % HAVO+
rob2_model <- lm(
  avg_advice ~ score_c * schoolweging_c + provider,
  data = model_data,
  weights = n_advised
)

# Without AMN and DOE (smallest providers)
rob3_model_data <- model_data |>
  filter(!provider %in% c("DOE", "AMN")) |>
  droplevels()

rob3_model <- lm(
  pct_havo_plus ~ score_c * schoolweging_c + provider,
  data = rob3_model_data,
  weights = n_advised
)

# "<5" replaced by 1 or 4 instead of 2.5: rebuild the data in a separate
# environment, so the main data and the saved .rds are not overwritten
fit_lt5 <- function(value) {
  env <- new.env()
  env$lt5_value <- value
  source(here("scripts", "03-clean-data.R"), local = env)
  data <- env$school_rq3 |>
    mutate(
      score_c = score - mean(score),
      schoolweging_c = schoolweging - mean(schoolweging),
      provider = relevel(factor(provider), ref = "LIB")
    )
  lm(
    pct_havo_plus ~ score_c * schoolweging_c + provider,
    data = data,
    weights = n_advised
  )
}

# Schoolweging effect and interaction (estimate and p-value) per model
get_effect <- function(model, interaction_term) {
  coefs <- coef(summary(model))
  tibble(
    "Schoolweging estimate" = coefs["schoolweging_c", "Estimate"],
    "Schoolweging p-value" = coefs["schoolweging_c", "Pr(>|t|)"],
    "Interaction estimate" = coefs[interaction_term, "Estimate"],
    "Interaction p-value" = coefs[interaction_term, "Pr(>|t|)"]
  )
}

robustness <- bind_rows(
  "Main model ('<5' = 2.5)" = get_effect(rq3_model, "score_c:schoolweging_c"),
  "Maths only" = get_effect(rob1_model, "maths_c:schoolweging_c"),
  "Average advice (1-6)" = get_effect(rob2_model, "score_c:schoolweging_c"),
  "Without AMN and DOE" = get_effect(rob3_model, "score_c:schoolweging_c"),
  "'<5' = 1" = get_effect(fit_lt5(1), "score_c:schoolweging_c"),
  "'<5' = 4" = get_effect(fit_lt5(4), "score_c:schoolweging_c"),
  .id = "Model"
)

knitr::kable(
  robustness,
  digits = c(0, 3, 4, 3, 4),
  caption = "Robustness checks: schoolweging effect and interaction per model"
)
```

The findings are robust to the main analytical choices. The negative schoolweging effect and the negative interaction remain significant when only maths results are used, when the two smallest providers are excluded and when advice is measured as an average on a 1-6 scale instead of a percentage. On that scale, a 10-point difference in schoolweging corresponds to about half a track lower advice. Include 2.5 vs 1 , 4 robustness check.

## Limitations and future work

However, certain limitations must be addressed. Since the data are school-level aggregates, the analysis cannot show whether individual pupils with the same result receive different advice. Also, the model is not causal as it shows that advice differs at equal results but not why. The gap could reflect teachers' prejudices, or information about pupils that the test does not capture. Add something about small schools??

Future work could use pupil-level data to understand why individual pupils with the same result receive different advice and also investigate teachers' advice processes to understand whether differences in advice stem from prejducices or from information beyond the test. It would also be interesting to examine schoolweiging in detail, for example, separating components such as parental education, migration background and neighborhood? Understanding these factors would help policy makers target interventions more precisely.

# AI use

The group used Clause Opus 5.5 and Clausde Sonnet 4.6. AI was used for technical tasks: R syntax, the data-inspection and data-cleaning code, and design suggestion for the final explanatory plot. It was also used to set up the report structure. The research question, hypotheses, choice of model, interpretation of the results and the content of the report were done without AI because… The AI worked from the assignment brief, a codebook, and our internally developed AI guidelines which governed how the AI…. and ensured accuracy, security, sovereignty and meaning in its outputs. The outputs were checked by peer review. These checks mattered because during the data cleaning phase, it left out the 2f language test result column. This was caught by a peer and was rectified before proceeding with the explanatory analysis.

# References

[Add references, e.g. DUO open data, the course data background page, R packages.]
